Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida

### 2022-01-28 revision/review_response.Rmd

---
output: 
  bookdown::word_document2:
    reference_docx: response_styles.docx
    number_sections: false
always_allow_html: true
urlcolor: blue
link-citations: true
bibliography: ../refs.bib
---

```{r setup, echo = F, warning = F, message = F, results = 'hide'}
# figure path, chunk options
knitr::opts_chunk$set(fig.path = 'figs/', warning = F, message = F, echo = F, cache = F, dev.args = list(family = 'serif'), dpi = 300, warning = F)
```

```{r echo = F, cache = F, eval = F}
spelling::spell_check_files('revision-response.Rmd')
```

We sincerely thank the reviewers for providing comments on our manuscript.  We provide a point-by-point response to these comments below. In general, most of the revisions address concerns raised by all three reviewers, including 1) improvement in writing style for clarity and brevity, and 2) additional information on sampling methods.  We have also addressed concerns of Reviewer 2 about potential nutrient cycling by adding a section in the discussion and concerns of Reviewer 4 regarding additional detail on the simulation modelling.

@@

Beck and colleagues report on the short- and mid-term effect of a very large nutrient-rich release from Piney Point. The authors focus on a few water quality indicators, but also include data for several additional responses (e.g., seagrass coverage). The results are largely what you would expect, initial rapid response of phytoplankton with cascading secondary responses.

* __Response__: We really appreciate your comments.  Please see the detailed responses below.

First, a few of the things that make this a worthwhile contribution to MPB. The authors should be commended for their quick response to data collection and the large number of data streams that were brought to the study. It really is quite impressive. Further, compared to citizen monitoring, I expect the data is high quality and consistent. Second, the writing was quite good; there were some sections where flow or organization could be improved (beginning of results; last section of discussion); however, these considerations are very minor.

* __Response__: Thank you for pointing out the merits of this paper and providing suggestions where the writing can be improved.  We have edited the sections you point out for brevity and clarity.

@@

Second, a few things that could be improved or warrant some critique. What is the rationale for the lack of continuous line numbers? Journal suggestion? Reviewing is so much easier when line numbers line up with the text and continue throughout the document. I realize this is not something that should be fixed now, but it is bothersome.

* __Response__: Unfortunately, the line numbers are generated automatically by the journal submission system and we have no control over how they are presented. We appreciate your use of both page and line numbers in your comments to reference text locations.

That said, my one major critique of the manuscript, one that is difficult to reconcile with the available data, is the speculative nature of last section of the Results, titled Potential nutrient cycling. This is clearly a very important section and, in my opinion, warrants additional consideration. The title of this section is correct, yes; thus, I would like to see 1) this section moved to the discussion, 2) the MAJOR lines of evidence supporting this thesis outlined and discussed consecutively in one section, 3) major alternatives entertained and addressed. The authors might also think about what it would take to make the connections more explicit, if possible (stable isotopes, modeling, etc.).

* __Response__: Thank you for this comment.  We agree this section is important as it relates the timeline of events in a plausible manner, although it is definitely speculative without more quantitative results linking the pieces.  As such, we have moved the content to the discussion and amended to text to include a more nuanced description of the trends, including what information may be needed to support the narrative.  This includes information from the final paragraph of the discussion, which was removed.  The new section is included from lines xxx to xxx.

@@

MINOR COMMENTS/SPECIFIC SUGGESTIONS

P3/L56: Recommend removing "in other countries". When I read that passage it seems to suggest a broader critique, one that is more political. That may not have been your intention, but that was how I read it.

* __Response__: Our point was to suggest that mining products in Florida are generally exported and have little benefit for the local communities, aside from localized economic benefits.  However, this may be a bit editorialized and was removed.

@@

P4/L21-24: I am not sure if "unanticipated releases" captures the spirt in which these releases occurred. In fact, the title of the FWC report was "Response of estuarine nekton to the regulated discharge of treated phosphate-production process water". This regulated discharge seemed to occur after extensive consultation with the EPA.

* __Response__: Agreed, we have removed "unanticipated". We agree that the releases often occur after consultation or approval from regulatory bodies, but they are generally unanticipated in that site management has not planned for these events, nor the potential impacts on the environment.

@@

P4/L51: worth indicating that these concentrations, especially for TP, are several orders of magnitude greater than typical surface waters?

* __Response__: The sentence was edited: "Water quality of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) were approximately three orders of magnitude higher than typical concentrations in Tampa Bay."

@@

P4/L56: deliberately "released"

* __Response__: Changed to "released under FDEP authorization".

@@

P5/L4: I see the use of the term lower Tampa Bay; how does that compare to Figure 1, the areas of interest? Ultimately, lower Tampa Bay is used extensively, but I don't see it defined. Also, is this region ecologically based or defined by currents?

* __Response__: The boundary between middle and lower Tampa Bay was added to Figure 1. These boundaries are used by the Tampa Bay Estuary Program to track annual attainment of water quality targets under a Reasonable Assurance plan for TMDL reporting to the Florida Department of Environmental Protection.  We do not define the rationale for the boundaries in the main text because the regulatory delineations of the bay are not of interest to a broader audience, but we use the delineation as it is useful for our discussion of trends in the paper.  Our area delineations (areas 1 - 3) were meant to provide a more relevant grouping for the paper.

P5/L22-27: I think a citation may be warranted here; while predictable, numerous studies ultimately supported the statement that nitrogen is limiting in Tampa Bay.

* __Response__: Agreed, we have revised this statement to include additional citations and explanation:

@@

     "Numerous studies, as well as the successful nutrient management paradigm for Tampa Bay, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14]. As such, we focus on nitrogen in our analyses as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production."

P5/L29: interesting that discharge was to Port Manatee and not Bishop Harbor like the release discussed previously.

* __Response__: Yes, we believe the intent was to minimize impacts to sensitive back-bay habitats that were observed in previous releases, although we have no information to verify this claim and have not included it in the text.

@@

P6/L24-39: that is an impressive and extensive list of agencies and partners; any potential drawbacks to so many data streams?

* __Response__: Data quality assurance and control was a primary concern with multiple partners collecting information.  We have included additional information about the data collection procedures and how the Tampa Bay Estuary Program worked to facilitate quality of the data.  Please see the additions starting on line xxx.

@@

P6/L54-65: is it routine to take water samples directly from the surface? Ultimately, were the authors worried about any potential stratification, particularly of Karenia brevis.

* __Response__:  Most of the water quality and phytoplankton samples were surface collections following standard sampling methods used by our partners (see the text addition about sampling).  Tampa Bay also has a very shallow average depth of just over 3m and is not strongly stratified for most of the year.  So, we feel confident that the majority of samples used in our analyses were generally representative of conditions in the water column. Anecdotally, benthic fishes were observed during the fish kills (e.g., eels, rays), so it is likely that *K. brevis* was present throughout the water column, although we have very little quantitative data to describe the vertical distribution.

@@

* __Response__: Seagrass transect monitoring in Tampa Bay has occurred annually since the mid-1990s using similar methods as in this paper [i.e., @Sherwood17].  The primary difference for the routine monitoring is the transect length, which is typically much longer than the 50m used in the 2021 rapid design. This shorter distance was chosen to allow for quicker sampling times and greater coverage in response to the rapidly changing conditions in 2021.  Thus, we consider the rapid design as providing data that are of similar quality to the long-term data, with notable exceptions indicated in the mansucript (i.e., no seasonal estimates).  Our rationale for evaluating Braun Blanquet coverage was to verify the results obtained evaluating frequency occurrence estimates.  Line xxx was modified to make this clear:

     "Tests using Braun Blanquet cover estimates confirmed the results from the frequency occurrence estimates (Tables S3, S4)."

P17/Potential nutrient cycling: Unfortunately, much of the latter portions of this section are speculative and better presented in the discussion. Further, it would have been nice to see something a bit more concrete linking the nutrient shift from photoplankton to macroalgae and release of macroalgae nutrients to K. brevis; stable isotopes perhaps; the nitrogen sources could have been highly enriched

* __Response__: This section was moved to the discussion, please see the response to your general comment above.

@@

* __Response__: We included some text describing this issue in the discussion.

     "Water quality conditions before and after passage of tropical storm Elsa may have also contributed to fish kills by reducing bottom-water dissolved oxygen. @Stevens06 documented impacts of a category 4 storm on fish resources in the Charlotte Harbor estuary, although tropical storm Elsa was much smaller and fish kills were documented prior to and after arrival of the storm.  Lack of continuous monitoring data for bottom waters in Tampa Bay prevents a more detailed assessment of impacts of the storm on water quality."

Discussion - In general, it would be useful if the team would spend some time establishing the prevalence of N-limitation in the system, either based on their own data or previous studies. The interpretation of plume impacts is predicated on the system being N-limited, so that needs to be established up front.

* __Response__: We agree that the nutrient limitation needs to be clearly stated as the basis for focusing primarily on nitrogen and not phosphorus. The second paragraph in the introduction was meant to provide this justification, but this was clearly insufficient as noted in the review.  We include some additional text and citations in the final paragraph of the introduction as further justification - see the response to the similar critique from reviewer 2.  Also note that phosphorus summary plots were previously included in the supplement.

### 2022-01-28 manu-draft.Rmd
getting through reviewer 2

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production in other countries [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed.  Effective management and final closure of these facilities are imperative to reduce threats to prior ecosystem recovery efforts and investments. The Piney Point facility located in Palmetto, Florida is a large, remnant gypstack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [see supplement for a history of the facility; @Henderson04].  Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.  Unanticipated releases from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those releases resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].

More recently, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.  In response, the Florida Department of Environmental Protection (FDEP) authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release water from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility. At that time, approximately 1.8 billion liters of mixed legacy phosphate mining wastewater and seawater from port dredging operations were being held in the failing gypstack. Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of both phosphorus and nitrogen.  Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) well outside normal ranges typical of surface waters in Tampa Bay. From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay.  Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].

@@

The above results can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April through September, starting with the influx of inorganic nitrogen from Piney Point. Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).  The initial peak in water quality parameters suggested a rapid response of the phytoplankton community as an increase in diatoms (e.g., centric species, such as *Skeletonema* sp., and also *Asterionellopsis* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were present in the initial discharge [@Bates76;@Domingues11].  Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.  However, filamentous cyanobacteria biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities. During peak macroalgae growth, TN and chl-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K. brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).  The co-occurring decline in macroalgae and increase in *K. brevis* suggests a release of nutrients from the former that could have stimulated growth of the latter, although residual water column nutrients from Piney Point may have also been present (as suggested by modelling efforts).  Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.  These distinct temporal periods were readily identified through an ordination plot (Figure S7) for the observed data in Figure \@ref(fig:nutrientflow).

@@

The observed conditions in Tampa Bay in 2021 following unanticipated releases from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.  These anomalous events included 1) a large diatom bloom in April in the vicinity of the release at Port Manatee, 2) high abundance of filamentous cyanobacteria in Anna Maria Sound and near Port Manatee, 3) medium to high bloom concentrations of the ride tide organism *K. brevis* in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.  The water quality conditions observed during the study period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).  The Piney Point event also represented an anomalous volume and load of labile nitrogen released directly into lower Tampa Bay.  Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous releases to Tampa Bay. An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [TBEP unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters released from Piney Point. Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].  External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).

@@

Limitations of our analyses are also important to note to inform future event-based monitoring and additional research.  All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and may not represent explicit cause and effect mechanisms.  However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.  Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.  Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the release, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.  Long-term fate of nutrients from Piney Point is uncertain and continued monitoring can further support understanding of ecosystem response in Tampa Bay beyond the initial results in 2021. Local, regional, and state partners should continue to pursue management and policy actions that can mitigate the continued threats of these facilities to the health of coastal resources.  These efforts are critical to managing Gulf of Mexico ecosystems given past successes and the need to address ongoing threats of climate change, human population growth, habitat loss, severe weather events, and recurring pollutant sources.

>>>>

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed.  Effective management and final closure of these facilities are imperative to reduce threats to prior ecosystem recovery efforts and investments. The Piney Point facility located in Palmetto, Florida is a large, remnant gypstack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [see supplement for a history of the facility; @Henderson04].  Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.  Releases from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those releases resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].

More recently, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.  In response, the Florida Department of Environmental Protection (FDEP) authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release water from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility. At that time, approximately 1.8 billion liters of mixed legacy phosphate mining wastewater and seawater from port dredging operations were being held in the failing gypstack. Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of both phosphorus and nitrogen.  Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) were approximately three orders of magnitude higher than typical concentrations in Tampa Bay. From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released under FDEP authorization to lower Tampa Bay.  Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].

@@

@@

The observed conditions in Tampa Bay in 2021 following releases from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.  These anomalous events included 1) a large diatom bloom in April in the vicinity of the release at Port Manatee, 2) high abundance of filamentous cyanobacteria in Anna Maria Sound and near Port Manatee, 3) medium to high bloom concentrations of the ride tide organism *K. brevis* in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.  The water quality conditions observed during the study period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).  The Piney Point event also represented an anomalous volume and load of labile nitrogen released directly into lower Tampa Bay.  Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous releases to Tampa Bay. An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [TBEP unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters released from Piney Point. Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].  External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).

The events of 2021 can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April through September, starting with the influx of inorganic nitrogen from Piney Point. Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).  The initial peak in water quality parameters suggested a rapid response of the phytoplankton community as an increase in diatoms (e.g., centric species, such as *Skeletonema* sp., and also *Asterionellopsis* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were present in the initial discharge [@Bates76;@Domingues11].  These results were evidenced by taxonomic enumeration of phytoplankton samples collected near Port Manatee.  Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.  However, filamentous cyanobacteria biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities. During peak macroalgae growth, TN and chl-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K. brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).  The co-occurring decline in macroalgae and increase in *K. brevis* suggests a release of nutrients from the former that could have stimulated growth of the latter.  Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.

Our quantitative results provide some evidence to support the progression of events outlined above as a flow of nutrients over time. The distinct temporal progression can be readily identified through an ordination plot (Figure S7) for the observed data in Figure \@ref(fig:nutrientflow).  Weekly summaries of the data are clearly separated in the ordination into monthly groups where different communities were dominant and is partially explained by orientation of the water quality vectors relative to cyanobacteria, diatoms, and *K. brevis*.  For example, total nitrogen and chlorophyll are strongly aligned with the *K. brevis* axis as nutrients were likely available in organic form during the peak of the red tide event.  However, this simple analysis only demonstrates an association in the observed data and cannot be verified without additional information.  Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.  Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the release, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.  Additional confounding variables can also obscure the association between water quality and community changes.  Bay conditions preceding the 2021 events, as well as the passage of tropical storm Elsa, could obscure these associations (described below).

## Additional interpretation of impacts

@@
